---
title: "2_Creating_and_Importing_Data"
author: "Matt Steele"
date: '2022-03-17'
output: html_document
---

===============================================================================

## Get and Set Working Directory ------------------------------------

```{r message=FALSE, warning=FALSE}

# get working directory

  getwd()

# set working directory
  
  setwd("~/RWorkshop Development/workshop_firstSteps_R")
  
# you can set wd by using menu > session > set working directory > choose directory


```


===============================================================================

## Part I - Creating Data in R | Vectors

    - one-dimensional sequence of data elements
    - use (c) To make a vector
    - separate elements using (,) (comma)
    - syntax == vectorSyntax <- c(object1, object2, object3)
    
```{r}

exampleVector <- c(1, 2, 3)
 exampleVector
str(exampleVector)

```

    - a vector can also be a string of factors or characters    
    
```{r}

wvTowns <- c("Morgantown", "Huntington", "Reedsville", "Charleston")
  print(c("My favorite town in West Virginia is ", sample(wvTowns, size = 1)))
  
```

===============================================================================

#3 Part II - Creating Data in R | Data Frames

    - two dimensional vectors that can multiple classes of values.
    - You can also create matrices in R. Matrices are similar to data frames but can also have a single data value
    - syntax == dataFrameSyntax <- data.frame(column1, column2, column3)

```{r}

  title <- c("Star Wars", "The Empire Strikes Back", "Return of the Jedi", "The Phantom Menace", "Attack of the Clones", "Revenge of the Sith", "Force Awakens", "The Last Jedi", "Rise of Skywalker")
  year <- c(1972, 1980, 1983, 1999, 2002, 2005, 2015, 2017, 2019)
  length.min <- c(121, 124, 133, 133, 142, 140, 135, 152, 141)
  box.office.mil <- c(787, 534, 572, 1027, 657, 849, 2059, 1333, 1074)
  
  starWars.data <- data.frame(title, year, length.min, box.office.mil)
  starWars.data

```

    # Function: str() -> will give you information of a variable
    
```{r}

help("str")

str(starWars.data)

```

    # Rename columns and tell R how to interpret the data frame
    
```{r}

starWars.data <- data.frame(Title = title, Year = year, Length = length.min, Gross = box.office.mil, stringsAsFactors = T)
  starWars.data

```

    
    # Subsetting Data Frames 
  
      - You can view and perform actions on columns using the ($) for subsetting

```{r}

  starWars.data$Gross
  summary(starWars.data$Gross) 


```

    # Mislabeled variables can be fixed
    
```{r}

starWars.data$Year <- as.character(starWars.data$Year)
  str(starWars.data)

```

    # Function: summary() -> will give you descriptive statistics on the data.frame
    
```{r}

help("summary")

summary(starWars.data)

```

    # Fix observations in a data frame
    
      - syntax == data.frame[rownumber, columnnumber] = newvalue
    
```{r}

  starWars.data [1, 2] = "1977"
    starWars.data  

```

    # New variable can be created using ($) and (=)
    
```{r}

starWars.data$Gross.billions = starWars.data$Gross / 1000

  starWars.data

```

    # Add new variables
    
```{r}

  mark <- c(37.5, 34.75, 34.25, 0, 0, .75, 0, 25.75, 7.5)
  carrie <- c(13.5, 22.75, 21.25, 0, 0, .5, 5.75, 12.75, 19.5)
  
  starWars.data$MarkScreenTime <- mark
  starWars.data$CarrieScreenTime <- carrie
  
  starWars.data

```


    # Add new observations

```{r}

help("rbind")

rogueOne <- data.frame(Title = "Rogue One", Year = 2016, Length = 133, Gross = 1051, Gross.billions = 1.051,  MarkScreenTime = 0, CarrieScreenTime = .25)

rbind(starWars.data, rogueOne)

```

===============================================================================

## Part IV: Import Data into R ------------------------------------------------


    # Function: read.*() ------------------------------------------------------

```{r}

help("read.table")

```

    # Function: read.table()
    
      - Allows you to add arguments and tell where seperations occur

```{r}

  my.house <- read.table("house_2018_bystate.csv", sep = ",", header = T)
    my.house
  
```

    # Function: read.csv()
    
      - Assumes comma seperated data

```{r}

  president.investigation <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/russia-investigation/russia-investigation.csv", stringsAsFactors = T)
    president.investigation

```

    # Package: foreign
    
      - Allows you to read proprietary software files (SPSS, SAS, STATA) 
      
```{r}

install.packages("foreign")

```

```{r}

library(foreign)

```

```{r}

help("read.spss")

 spssDemo <- read.spss("demo.sav", use.value.labels = T)
    view(spssDemo)
    

```

==============================================================================

## Part V - Viewing your data


    # Function: view() - view the full data frame
    
```{r}

help("View")

  view(president.investigation)

```
  
  
    # Function: nrow() - view the number of rows/observations
  
```{r}

help("nrow")

  nrow(president.investigation)

```

    # Function: ncol() - view the number of columns/variables
    
```{r}

help("ncol")

  ncol(president.investigation)

```
  
    # Function: colnames() - view the names of columns/variables
    
```{r}

help("colnames")

  colnames(president.investigation)

```

    # Function: str() - view the structure of data
    
```{r}

help("str")

  str(president.investigation)

```

    # Function: summary() - basic descriptive statistics
    
```{r}

help("summary")

  summary(president.investigation)

```



==============================================================================

## Part VI - Exporting your data

    # Function: write.*
    
```{r}

help("write.table")

  write.csv(president.investigation, "President_investigation.csv")

```
    


    