---
title: "3_Introduction_to_Tidyverse_Package"
author: "Matt Steele"
date: '2022-03-17'
output: html_document
---


===============================================================================

## Get and Set Working Directory ------------------------------------

```{r message=FALSE, warning=FALSE}

# get working directory

  getwd()

# set working directory
  
  setwd("~/RWorkshop Development/workshop_firstSteps_R")
  
# you can set wd by using menu > session > set working directory > choose directory


```

=========================================================================

## Packages -----------------------------------------------------------

    ## Packages that you will need to have installed for this workshop

```{r message=FALSE, warning=FALSE}

install.packages("tidyverse")

```

    ## Packages that you will need to load for this workshop

```{r message=FALSE, warning=FALSE}

library(tidyverse)

```

=============================================================================

## Resources ---------------------------------------------------------------

    Tidyverse Documentation - https://www.tidyverse.org/
    
    RStudio Cheatsheets - https://www.rstudio.com/resources/cheatsheets/ 
    
===============================================================================

## Tidyverse ----------------------------------------------------------------

    # collection of packages for data analysis and data visualizations

      ## ggplot2 - data visualization 
          - https://ggplot2.tidyverse.org/
      
      ## tibble - lighter and more user-friendly version of data.frame() 
          - https://tibble.tidyverse.org/
      
      ## tidyr - create tidy and meaningfully arranged data 
          - https://tidyr.tidyverse.org/
      
      ## readr - sidesteps the limitations of the R functions when importing data into R 
          - https://readr.tidyverse.org/
      
      ## purrr - better functional programming 
          - https://purrr.tidyverse.org/
      
      ## dplyr - data manipulation tools 
          - https://dplyr.tidyverse.org/


==============================================================================

## Pipe Operator ( %>% ) -----------------------------------------------------

    # allows you to separate and unite individual operations on an object
        # CTRL + SHIFT + M (windows)
        # CMD + SHIFT + M (mac)
        
    # data %>% operation A %>% operation B
        
```{r}

my.house %>%
    group_by(Majority_votes) %>%
    summarise(count = n(), avg.vote = mean(Seat_difference, na.rm = T)) %>%
    filter(count >1)

```

===============================================================================

## Part I - Read data into R using Tibble ------------------------------------

    # allows for lighter and cleaner data frames
    
```{r}
help('as_tibble')

```

```{r}

my.house <- read.csv("house_2018_bystate.csv", stringsAsFactors = T)

my.house <- as_tibble(my.house)

  summary(my.house)

```

    # Function: read_*()

```{r}

help("read_csv")

```

```{r}

sports.donations <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/sports-political-donations/sports-political-donations.csv")

  sports.donations

```

===============================================================================

## Part II - Manipulate your data with the Dplyr Package ----------------------


    # Function: filter() - subsets data according to criteria
    
```{r}

?dplyr::filter

```
    
```{r}

# Filter data to just include observations where the investigation variable is russia

filter(president.investigation, investigation == "russia")

# Add boolean operators (| == or) and (& == and)

sports.donations
sports.donations$Amount <-  parse_number(sports.donations$Amount)

republican.donations <- filter(sports.donations, Party == "Republican" | Party == "Bipartisan")
  republican.donations
  mean(republican.donations$Amount)

```

    # Function: select() - keeps ONLY selected variables
    
```{r}

?dplyr::select

```

```{r}

majority.seats <- select(my.house, Majority_seats, Vote_difference, Seat_difference)
  summary(majority.seats)


```

    # Function: mutate() - create a new variable and append it to the data frame
    
```{r}

?dplyr::mutate

```

```{r}

my.house <- mutate(my.house, Seat_diff = abs(Republican_seats - Democrat_seats))
    my.house

```

    # Function: transmute() - creates a new variable and discards the rest of the data frame
    
```{r}

?dplyr::transmute

```

```{r}

seat.difference <- transmute(my.house, Seat_diff = abs(Republican_seats - Democrat_seats))
  seat.difference

```

    # Function: arrange() - sort data frame by variables
    
```{r}

?dplyr::arrange

```
    
```{r}

arrange(sports.donations, +(Amount))
arrange(sports.donations, -(Amount))

```


    # Function: summarise() - get a summary of a group
    
```{r}

?dplyr::summarise

```
    
```{r}

dplyr::summarise(republican.donations, avgAmount = mean(Amount)) 

```


    # Function: group_by() - group a categorical variable and create a new table
 
 
```{r}

?dplyr::group_by

```
    
```{r}

republican.byOwner <- republican.donations %>% 
  group_by(Owner) %>%
  summarise(avg.amount = mean(Amount))

  republican.byOwner


```

    # Function: sample() - randomly sample from the data frame
    
```{r}

?dplyr::sample_n

?dplyr::sample_frac

```


```{r}

sample.donations <- sports.donations %>%  sample_n(20, replace = T)
  sample.donations
  
sample.donations <- sports.donations %>% sample_frac(.25, replace = T)
  sample.donations

```

===============================================================================

## Part III - Tidy up your Data with the Tidyr package -----------------------

    # Function: pivot_longer() - convert variables to observations

```{r}

?tidyr::pivot_longer

```


```{r}

covid19.us <- read.csv("time_series_covid19_confirmed_US.csv", stringsAsFactors = F)
      covid19.us <- as_tibble(covid19.us)
      covid19.us

```


```{r}

covid.clean <- covid19.us %>% pivot_longer(January.22..2020:October.19..2020, values_to = "Cases", names_to = "Date") %>% arrange(-Cases)

  covid.clean

```

    # Function: as.Date() - convert =value to a date value
    
```{r}

?as.Date

```

```{r}

covid.clean$Date <- as.Date(x = covid.clean$Date, tryFormats = c("%B.%d..%Y"))
      covid.clean

```



    # Function: separate() - create new variables by separating elements from a variable
    
```{r}

?tidyr::separate

```

```{r}

covid.clean <- covid.clean %>% 
  separate(Combined_Key, sep = ", ", into = c("City", "State", "Country")) %>% 
  separate(Date, sep = "-", into = c("Year", "Month", "Day"))

  covid.clean

```

    # Function: unite() - create a new variable by uniting element from multiple variables
    
```{r}

covid.cleanUnited <- covid.clean %>% 
  unite("Date", sep = "-", c("Year", "Month", 'Day'))

covid.cleanUnited$Date <- as.Date(covid.cleanUnited$Date)

  covid.cleanUnited

```

covid.united <- covid.separate.date %>% unite("Date", sep = "-", c("Year", "Month", 'Day'))
      covid.united$Date <- as.Date(covid.united$Date)

==============================================================================
    
## Part IV - Visualize your data with GGplot2

      # Let create a bar graph using cases by state in October

```{r}

covid.cleanOct <- covid.clean %>% filter(Month == 10) %>% na.omit # only include October

```

      # Set the parameters
      
      
```{r}

bar.covid.state <- ggplot(covid.clean, aes(State), (x=factor(trunc(runif(10000, 0, 100)), ordered=T)))
      bar.covid.state

```

      # Plot the data
      
```{r}

bar.covid.state + geom_bar(fill = "coral") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "COVID Cases by State", 
       subtitle = "October 2020", 
        x = NULL, 
        y = NULL)

```


      # Let's create a line graph of cases over time
      
```{r}


line.covid.time <- ggplot(covid.cleanUnited, aes(x = Date, y = Cases)) 
      

line.covid.time + stat_smooth(show.legend = F, color = "coral", n = 180) + 
  theme_minimal() +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs( x = NULL, 
        y = NULL,
        title = "COVID-19 Cases in the United States",
        subtitle = "through October 2020"
    )

```



  